---
title: "MATH 216 Homework 4"
author: "Jacob Dixon"
output: html_document
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# For data manipulation and visualization
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(knitr))
# For US county and state maps
suppressPackageStartupMessages(library(maps))
# For loading in shapefiles
suppressPackageStartupMessages(library(rgdal))
suppressPackageStartupMessages(library(maptools))
# For interactive maps
suppressPackageStartupMessages(library(leaflet))
```

## Admistrative:

Please indicate

* Who you collaborated with:Alison Cook
* Roughly how much time you spent on this HW: 8h
* What gave you the most trouble: Lots of little things, cleaning up text, issues with overwriting files, leaflet.
* Any comments you have: I am worried that spending this much time on these assignments will give me very little time to work on the independent project.





## Question 1:

```{r, cache=TRUE, echo=FALSE}
# Load state and county map of US in 2010 from the maps package and convert them
# to data frames so that we can ggplot them.
US_state <- map_data("state") %>% 
  tbl_df()
US_county <- map_data("county") %>% 
  tbl_df()
voter_data <- read.csv("COUNTY.csv", header=TRUE) %>% tbl_df()

shapefile_name <- paste(getwd(), "/VT_census_tracts/tl_2015_50_tract.shp", sep="")
VT <- readOGR(shapefile_name, layer = "tl_2015_50_tract", verbose = FALSE)
```

### Choropleth Map of US Voter Behavior in 2000

Download the results of the 2000 election from the [School of Public Affairs at 
American University in DC](http://www.american.edu/spa/ccps/Data-Sets.cfm) and 
create a map involving only the lower 48 states that show voter behavior at a 
**county** level. To keep things simple let's only consider George W. Bush, Al
Gore, and Ralph Nader. Your write-up must include:

1. A [choropleth map](https://en.wikipedia.org/wiki/Choropleth_map) where
    * Counties are filled with red when they tend to favor Bush
    * Counties are filled with white when they tend to be split
    * Counties are filled with blue when they tend to favor Gore and Nader
2. An answer to the following question: which states exhibit the greatest
**within state heterogeneity** in voting? Come up with a mathematical
justification.

### Write-Up

```{r, fig.width=12, fig.height=6}
# This function eliminates all non-alphanumeric characters and spaces and converts
# all text to lower case:
clean_text <- function(text){
  text <- gsub("[^[:alnum:]]", "", text)
  text <- gsub(" ", "", text)
  text <- tolower(text)
  return(text)
}

#Clean up the US_county dataset using the clean_text function. Gets rid of spaces
#in state names and any potential capitalzation 
US_county$region <- clean_text(US_county$region)
US_county$subregion <- clean_text(US_county$subregion)

#clean up voter_data for the same things as above
voter_data$STATE <- clean_text(voter_data$STATE)
voter_data$COUNTY <- clean_text(voter_data$COUNTY)

#Join the two datasets by state and county
location_and_vote <- left_join(US_county, voter_data, by = c( "region" = "STATE", 
                                                              "subregion" = "COUNTY"))
          
#Make them not factors, so that my function below works
location_and_vote$PBUSH <- as.numeric(as.character(location_and_vote$PBUSH))
location_and_vote$PGORE <- as.numeric(as.character(location_and_vote$PGORE))
location_and_vote$PNADER <- as.numeric(as.character(location_and_vote$PNADER))


#Assign areas to those breakdowns 
location_and_vote$favor <-  ifelse(location_and_vote$PBUSH >= 0.5, -1,
                                   ifelse(location_and_vote$PGORE + 
                                            location_and_vote$PNADER >= 0.5, 1, 0))



#Map that distribution 
ggplot(data=NULL) +
  geom_polygon(data=location_and_vote, aes(x=long, y=lat, group=group, fill=favor)) +
  geom_path(data=location_and_vote, aes(x=long, y=lat, group=group), col="black", size=0.05) +
  geom_path(data=US_state, aes(x=long, y=lat, group=group), col="black", size=0.1) +
  scale_fill_gradient2(name = "", low ="red", high ="blue", mid ="white") +
  coord_map()


#Determine the number of counties for each state 
county_count <- US_county %>% distinct(subregion) %>%
  group_by(region) %>% tally() %>% rename(count = n)

#Aggregate and sum, ignoring NAs, grouping by region 
election_score_state <- aggregate(location_and_vote$favor, 
                                  by=list(region=location_and_vote$region), 
                                  FUN=sum, na.rm = TRUE)

#Creat a proportion of counties 
election_score_state <- election_score_state %>% mutate(prop = x/county_count$count)

#Corrects for any negative values
election_score_state$prop <- abs(election_score_state$prop)

#Arrange by which have the most heterogeneity
election_score_state <- election_score_state %>% arrange(prop)

#Print out a table of that descending list - looks like no data for DC, they are probably mad
#because of it...can't make their own laws and what not 
kable(election_score_state)

```






## Question 2:

In this question, you must make an interactive "Single File" Shiny app that uses
Leaflet. For all 184 census tracts in VT in the 2010 census, present
information on the proportion of the population that is either

* White
* African-American
* Hispanic
* Asian and Pacific Island
* Other, including those of two or more races

Use [Social Explorer](http://www.socialexplorer.com/) to get census data. I did
a demo of this in class. If you don't remember how or are stuck, please speak to
me or get help from your peers. Do not submit copies of the same file.

There should be some mechanism in your Shiny app that allows one the user to toggle
between the different ethnic groups.


### Loading Shapefile Data

Here is some starter code:

```{r}
#Import the social explorer data
census_2010 <- read.csv("VT_COUNTY_SE.csv", header=TRUE) %>% tbl_df()

#Rename the columns in the table to be more clear to me 
census_2010 <- rename(census_2010, total_pop = `SE_T055_001`, white_alone = `SE_T054_002`, 
                      black_alone = `SE_T054_003`,asian_alone = `SE_T054_005`, 
                      hawaii_pac_alone = `SE_T054_006`, other_alone = `SE_T054_007`, 
                      two_more = `SE_T054_008`, hisp_latino = `SE_T055_010`)


                      
#Select out only those that I am interested in 
census_2010_race <- select(census_2010, Geo_NAME, Geo_QName, Geo_FIPS, total_pop, white_alone, 
                           black_alone, asian_alone, hawaii_pac_alone, other_alone, two_more, 
                           hisp_latino)


#Calculate the proportions and do some combining to meet the groupings in the question 
census_2010_race$white_prop <- census_2010_race$white_alone/census_2010_race$total_pop
census_2010_race$black_prop <- census_2010_race$black_alone/census_2010_race$total_pop
census_2010_race$hispanic_prop <- census_2010_race$hisp_latino/census_2010_race$total_pop
census_2010_race$apac_prop <- census_2010_race$hawaii_pac_alone + census_2010_race$asian_alone/
  census_2010_race$total_pop
census_2010_race$other_prop <- census_2010_race$other_alone + census_2010_race$two_more/
  census_2010_race$total_pop


#Do the same thing we did for map_df. 
VT$id <- rownames(VT@data)
VT_points <- fortify(VT, region="id")
VT_df <- inner_join(VT_points, VT@data, by="id")

#Join race data and vermont_df 
VT_df$GEOID <- as.numeric(as.character(VT_df$GEOID))
VT_census <- left_join(VT_df, census_2010_race, by = c("GEOID" = "Geo_FIPS"))



#make some maps, first, Hispanic 
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = colorBin("Blues", VT_census$hispanic_prop)(VT_census$hispanic_prop)
  )

#White
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = colorBin("Blues", VT_census$white_prop)(VT_census$white_prop)
  )


#African American
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = colorBin("Blues", VT_census$black_prop)(VT_census$black_prop)
  )


#Asian and Pacific Island
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = colorBin("Blues", VT_census$apac_prop)(VT_census$apac_prop)
  )



#Other, including those of two or more races
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = colorBin("Blues", VT_census$other_prop)(VT_census$other_prop)
  )




```


### Write-Up

Upload your shiny app to the Middlebury Shiny Server (see Lecture 16) and post
the url to the app [here](https://www.nhl.com/), replacing the nhl.com link with 
the link to your app.

Comment on general ethnic demographic trends that's more substantive than just
"Vermont is really white."



