---
title: "MATH 216 Homework 4"
author: "Jacob Dixon"
output: html_document
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# For data manipulation and visualization
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(knitr))
# For US county and state maps
suppressPackageStartupMessages(library(maps))
# For loading in shapefiles
suppressPackageStartupMessages(library(rgdal))
suppressPackageStartupMessages(library(maptools))
# For interactive maps
suppressPackageStartupMessages(library(leaflet))
```

## Admistrative:

Please indicate

* Who you collaborated with:Alison Cook
* Roughly how much time you spent on this HW: 10h
* What gave you the most trouble: Leaflet is difficult, because it is different 
* Any comments you have: Still not happy with how the distribution maps look, but I don't know how to improve them. 





## Question 1:

```{r, cache=TRUE, echo=FALSE}
# Load state and county map of US in 2010 from the maps package and convert them
# to data frames so that we can ggplot them.
US_state <- map_data("state") %>% 
  tbl_df()
US_county <- map_data("county") %>% 
  tbl_df()
voter_data <- read.csv("COUNTY.csv", header=TRUE) %>% tbl_df()

shapefile_name <- paste(getwd(), "/VT_census_tracts/tl_2015_50_tract.shp", sep="")
VT <- readOGR(shapefile_name, layer = "tl_2015_50_tract", verbose = FALSE)
```

### Choropleth Map of US Voter Behavior in 2000

Download the results of the 2000 election from the [School of Public Affairs at 
American University in DC](http://www.american.edu/spa/ccps/Data-Sets.cfm) and 
create a map involving only the lower 48 states that show voter behavior at a 
**county** level. To keep things simple let's only consider George W. Bush, Al
Gore, and Ralph Nader. Your write-up must include:

1. A [choropleth map](https://en.wikipedia.org/wiki/Choropleth_map) where
    * Counties are filled with red when they tend to favor Bush
    * Counties are filled with white when they tend to be split
    * Counties are filled with blue when they tend to favor Gore and Nader
2. An answer to the following question: which states exhibit the greatest
**within state heterogeneity** in voting? Come up with a mathematical
justification.

### Write-Up

```{r, fig.width=12, fig.height=6, echo=FALSE}
# This function eliminates all non-alphanumeric characters and spaces and converts
# all text to lower case:
clean_text <- function(text){
  text <- gsub("[^[:alnum:]]", "", text)
  text <- gsub(" ", "", text)
  text <- tolower(text)
  return(text)
}

#Clean up the US_county dataset using the clean_text function. Gets rid of spaces
#in state names and any potential capitalzation 
US_county$region <- clean_text(US_county$region)
US_county$subregion <- clean_text(US_county$subregion)

#clean up voter_data for the same things as above
voter_data$STATE <- clean_text(voter_data$STATE)
voter_data$COUNTY <- clean_text(voter_data$COUNTY)

#Join the two datasets by state and county
location_and_vote <- left_join(US_county, voter_data, by = c( "region" = "STATE", 
                                                              "subregion" = "COUNTY"))
          
#Make them not factors, so that my function below works
location_and_vote$PBUSH <- as.numeric(as.character(location_and_vote$PBUSH))
location_and_vote$PGORE <- as.numeric(as.character(location_and_vote$PGORE))
location_and_vote$PNADER <- as.numeric(as.character(location_and_vote$PNADER))


#Assign areas to those breakdowns 
location_and_vote$favor <-  ifelse(location_and_vote$PBUSH >= 0.5, -location_and_vote$PBUSH,
                                   ifelse(location_and_vote$PGORE + 
                                            location_and_vote$PNADER >= 0.5, 
                                          location_and_vote$PGORE + 
                                            location_and_vote$PNADER, 0))



#Map that distribution 
ggplot(data=NULL) +
  geom_polygon(data=location_and_vote, aes(x=long, y=lat, group=group, fill=favor)) +
  geom_path(data=location_and_vote, aes(x=long, y=lat, group=group), col="black", size=0.05) +
  geom_path(data=US_state, aes(x=long, y=lat, group=group), col="black", size=0.1) +
  scale_fill_gradient2(name = "Party Votes", low ="red", high ="blue", 
                       mid ="white", breaks = c(0.75, 0.0, -0.75), 
                       labels = c("Gore & Nader", "Split", "Bush")) +
  coord_map()


#Determine the number of counties for each state 
county_count <- US_county %>% distinct(subregion) %>%
  group_by(region) %>% tally() %>% rename(count = n)

#Aggregate and sum, ignoring NAs, grouping by region 
election_score_state <- aggregate(location_and_vote$favor, 
                                  by=list(region=location_and_vote$region), FUN=sum, na.rm = TRUE)

#Creat a proportion of counties 
election_score_state <- election_score_state %>% mutate(prop = x/county_count$count)

#Corrects for any negative values
election_score_state$prop <- abs(election_score_state$prop)

#Arrange by which have the most heterogeneity
election_score_state <- election_score_state %>% arrange(prop)

#Print out a table of that descending list - looks like no data for DC, they are probably mad
#because of it...can't make their own laws and what not 
kable(election_score_state)

```

The choropleth map above shows US voter behaviour in the 2000 presidential election. Blue states favored Gore and Nader, while red favored Bush, with white being relatively split (it should be noted that several are gray, these represent no data, likely coming from restructuring of counties since the election took place). I think it is interesting to see how few counties were split in their choice for president. 

The table below the figure is a ranking of states from most heterogenous to least. This metric takes the proportion of voters, by county, voting for members of each party and combines them. This would result in values closest to 0 being most split. This was calculated by dividing the summed proportion of votes in each county by the number of counties. This tells you how heterogeneous the counties are. 




## Question 2:

In this question, you must make an interactive "Single File" Shiny app that uses
Leaflet. For all 184 census tracts in VT in the 2010 census, present
information on the proportion of the population that is either

* White
* African-American
* Hispanic
* Asian and Pacific Island
* Other, including those of two or more races

Use [Social Explorer](http://www.socialexplorer.com/) to get census data. I did
a demo of this in class. If you don't remember how or are stuck, please speak to
me or get help from your peers. Do not submit copies of the same file.

There should be some mechanism in your Shiny app that allows one the user to toggle
between the different ethnic groups.


### Loading Shapefile Data

Here is some starter code:

```{r, fig.width=12, fig.height=6, echo=FALSE}
#Import the social explorer data
census_2010 <- read.csv("VT_COUNTY_SE.csv", header=TRUE) %>% tbl_df()

#Rename the columns in the table to be more clear to me 
census_2010 <- rename(census_2010, total_pop = `SE_T055_001`, white_alone = `SE_T054_002`, 
                      black_alone = `SE_T054_003`,asian_alone = `SE_T054_005`, 
                      hawaii_pac_alone = `SE_T054_006`, other_alone = `SE_T054_007`, 
                      two_more = `SE_T054_008`, hisp_latino = `SE_T055_010`)


                      
#Select out only those that I am interested in 
census_2010_race <- select(census_2010, Geo_NAME, Geo_QName, Geo_FIPS, Geo_TRACT, total_pop, 
                           white_alone, 
                           black_alone, asian_alone, hawaii_pac_alone, other_alone, two_more, 
                           hisp_latino)


#Calculate the proportions and do some combining to meet the groupings in the question 
census_2010_race$white_prop <- census_2010_race$white_alone/census_2010_race$total_pop
census_2010_race$black_prop <- census_2010_race$black_alone/census_2010_race$total_pop
census_2010_race$hispanic_prop <- census_2010_race$hisp_latino/census_2010_race$total_pop
census_2010_race$apac_prop <- census_2010_race$hawaii_pac_alone + census_2010_race$asian_alone/
  census_2010_race$total_pop
census_2010_race$other_prop <- census_2010_race$other_alone + census_2010_race$two_more/
  census_2010_race$total_pop



#make some maps, first, Hispanic 
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="bl", weight=1) %>% 
  addLegend("bottomright", values = ~census_2010_race$hispanic_prop,
            title = "Proportion Hispanic",
            labFormat = labelFormat(prefix = ""),
            opacity = 1, pal = colorNumeric("Blues", census_2010_race$hispanic_prop)) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = ~colorQuantile("Blues", 
                           census_2010_race$hispanic_prop)(census_2010_race$hispanic_prop)
  )

#White
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addLegend("bottomright", values = ~census_2010_race$white_prop,
            title = "Proportion White",
            labFormat = labelFormat(prefix = ""),
            opacity = 1, pal = colorNumeric("Blues", census_2010_race$white_prop)) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = ~colorQuantile("Blues", 
                           census_2010_race$white_prop)(census_2010_race$white_prop)
  )


#African American
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addLegend("bottomright", values = ~census_2010_race$black_prop,
            title = "Proportion African American",
            labFormat = labelFormat(prefix = ""),
            opacity = 1, pal = colorNumeric("Blues", census_2010_race$black_prop)) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = ~colorQuantile("Blues", 
                           census_2010_race$black_prop)(census_2010_race$black_prop)
  )


#Asian and Pacific Island
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addLegend("bottomright", values = ~census_2010_race$apac_prop,
            title = "Proportion Asian & P. Islander",
            labFormat = labelFormat(prefix = ""),
            opacity = 1, pal = colorNumeric("Blues", census_2010_race$apac_prop)) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = ~colorQuantile("Blues", 
                           census_2010_race$apac_prop)(census_2010_race$apac_prop)
  )

#Other, including those of two or more races
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addLegend("bottomright", values = ~census_2010_race$other_prop,
            title = "Proportion Other",
            labFormat = labelFormat(prefix = ""),
            opacity = 1, pal = colorNumeric("Blues", census_2010_race$other_prop)) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = ~colorQuantile("Blues", 
                           census_2010_race$other_prop)(census_2010_race$other_prop)
  )



```

The maps above show the proportions of various races in Vermont, based on census information. While Vermont is the second whitest state in the union, it does have minority populations within its borders. Looking at the maps above, one can see minority populations are most dense near Vermont's cities of Burlington, Rutland, and the capital at Montpelier. As much of Vermont is now farmland because of the mass logging that took place two centuries ago, most of the state is rural. This combined with its geographic location in the north and lack of many large cities has contributed to its limited racial diversity. 

### Write-Up

Upload your shiny app to the Middlebury Shiny Server (see Lecture 16) and post
the url to the app [here](https://www.nhl.com/), replacing the nhl.com link with 
the link to your app.

Comment on general ethnic demographic trends that's more substantive than just
"Vermont is really white."



